Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-22 src/tex/ms.tex
Finished section 2

\begin{abstract}
	We investigate the possibility of improving the accuracy of a
	phenomenological waveform model, IMRPhenomD, by jointly optimizing all the
	calibration coefficients at once, given a set of numerical relativity (NR)
	waveforms. When IMRPhenomD was first calibrated to NR waveforms, different
	parts of the waveform were calibrated separately. Using \ripple, a library
	of waveform models compatible with automatic differentiation, we can perform
	gradients-based optimization to all the waveform coefficients the same time,
	which should improve the quality of waveform by capturing correlations
	between previous separately fitted parts. We found that after recalibration,
	the median mismatch between the model and NR waveforms decrease by $50\%$.
	We further explore how different parts of the source parameter space respond
	to the optimization procedure. We find that the degree of improvement
	correlates with the spins of the source. This work shows a promising avenue
	to help understand and treat systematic error in waveform models.
\end{abstract}

@@

% Brief intro of waveform models

Many data analysis tasks in today's gravitational wave (GW) astrophysics, such
as match filtering search \citep{fill} and parameter estimation\citep{fill}, rely on accurate waveform
models. Because using numerical relativity (NR) waveforms in these tasks is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants, which are effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, Numerical Relativity (NR)
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2}
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detail of construction of these models are different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of a
waveform model is determined by the ansatz used and the accuracy of the
calibrated parameters.

% Why there is a need

The LIGO-VIRGO-KAGRA (LVK) collaboration
(\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym}.) has recently started their fourth observational
run on May 26, 2023. Because of the sensitivity improvement, the new run is
expected to double the size of current binary black hole (BBH) observations
\citep{abbott2020prospects}. The improved sensitivity also implies we expect to
detect individual events with a higher signal-to-noise ratio (SNR). This means
we can resolve more features in the signal, at the same time put a more
stringent requirement on the accuracy of our waveform model\citep{fill}.

% AD waveforms

Because of the large number of calibration parameters (often in a couple hundreds if not
more), waveform models is usually calibrated in pieces \citep{fill}. This
ignores the correlation between different parts of the waveform model and limits
it quality. Recently, there has been efforts in rebuilding waveform models
\citep{fill} that supports automatic differentiation (AD)\citep{fill}, which is
a technique used to compute derivatives of functions down to machine precision
without issues of scaling up to high dimension or expression swelling. In
particular, {\ripple} \citep{ripple} exposes the calibration parameters to the
user. This allows us to make use of infrastructures that are heavily used in
machine learning, such as gradient descent and back propagation\cite{fill}, to
improve the calibration of the waveform models.

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD\citep{fill}, by jointly optimizing all the
calibration coefficients given a set of NR waveforms. Training on a few NR
waveforms, we demonstrate one can improve the match between IMRPhenomD and NR
waveforms over a decently sized parameter space, up to mass ratio $q=8$ We also
explore how different parts of the source parameter space (e.g. the primary and
second spins) respond to the optimization procedure, by optimizing the waveform
separately for different regions of the parameter space. This gives us hints on
whether the waveform model performs equally well in different regions of the
parameter space.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of optimized waveforms with original waveforms. We also show how the
optimization result differs with waveforms of different source parameters. In
Sec.~\ref{sec:discussion}, we address the difference between our calibrating
procedure with \citep{khan2016frequency}. We also explain how reduced spin
parameterization affects the accuracy of the model.

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), $f_{\mathrm{max}}$ and $f_{\mathrm{min}}$ are
the relevant maximum and minimum frequencies for the integration.
We note here that the mismatch can be viewed as the mean square error (MSE) between the
two waveforms.

\te{Up to here}

Since we wish to optimize the model over the whole parameter space, we need to compare multiple model generated waveforms with NR waveforms. However, mismatch only quantifies the difference between IMRPhenomD and NR waveform for one particular set of intrinsic parameters. To take into account of various different waveforms in the parameter space, we pick waveforms from the different parts of the parameter space. We define the loss function as an average of training waveforms in two ways, the simple average of mismatches and the normalize average of mismatches,  
\begin{align}
	\mathcal{L}_{\mathrm{mean}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{fl}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of individual training waveforms.
Note that we choose to use two different averages, since they have different 
preferences in optimization base on waveform mismatches. For the first choice, 
simple average serves as the simplest choice of loss function, but is prone to 
be dominated by a single waveform with a large mismatch. Other waveforms with 
smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such optimization. Alternatively, the second choice, normalized 
average eliminates the aforementioned issue. Nevertheless, it excludes the
information on initial mismatches. $\mathcal{L}_{\mathrm{fl}}$ restricts every
training waveform to decrease at similar rates, hence it is hard to obtain
optimized waveforms with mismatches in the same order of magnitude. Instead,
their ratios in mismatches would remain approximately the same. Conversely,
$\mathcal{L}_{\mathrm{mean}}$ allows the loss function to automatically adjust
and individual mismatches would be in a similar order of magnitude after 
optimization. In this paper, we showcase the results of using both loss functions 
and examine the differences between them.

@@

To compute the loss functions, we have to take NR waveforms for calculating the mismatch. We choose 11-16 NR waveforms from the set of waveforms used in the original
calibration process as training waveforms. Originally, 19 waveforms are taken from 
NR simulations for calibrating IMRPhenomD \citep{khan2016frequency}, 
which are waveforms from the SXS catalog \citep{boyle2019sxs} or BAM simulation. 
As BAM waveforms are not publicly available, we cannot take the identical training 
set as them. Instead, we take the available waveforms from the SXS catalog to construct our loss function. Training waveforms used are listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. The training waveforms chosen has maximum mass ratio to be 8. This is because SXS catalog does not have NR waveforms with extremely high mass ratio. In fact, the SXS catalog only has NR waveforms with $q\leq10$. Nevertheless, we are interested in the behavior of IMRPhenomD model with small $q$, as most BBH events observed from LIGO have $q\leq8$. Hence, we calibrate IMRPhenomD with waveforms of $q\leq8$.

In the SXS catalog, NR waveforms are in the form of time-series strain. Since time-series data is oscillatory, performing optimization in the time-domain is not ideal. Hence, we transform NR waveforms to frequency-domain to compare with IMRPhenomD waveforms with the
same intrinsic parameters. We taper the time-series using Tukey window.
\footnote{ Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown and $T$ is the duration of the
	entire GW strain. } 
Then, the frequency spectra can be obtained by taking the Fourier transform of the time-series.

%\te{I think this discussion needs to be much more precise. Having a waveform model calibrated to particular noise curve could make sense, but its inherently different to just fitting the phase and the amplitude separately like they do in the original paper. So I would recommend that our default is to use a flat PSD and then in addition discuss the one with the PSD. This way, I think we demonstrate more clearly that its the high dimensional fitting that helps, not the just changing the metric used when fitting.}
Other than NR waveforms, one need to choose a relevant noise PSD for 
mismatch. We have opted to use a flat PSD for this purpose, as it provides results that are independent of the detector sensitivity and mass scale. The use of a flat PSD ensures that the improvement in accuracy is due mainly to the difference in high-dimensional fitting and piece-wise fitting, but not due to the use of a different mass scale. Furthermore, we are interested in examining the effect of introducing a detector PSD on the optimization process. For this, we have chosen the zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}. Since the total mass of the system scales with the frequency of the waveform, we must choose a corresponding mass scale to match the frequency range of our noise PSD. We selected an arbitrary mass scale of $M=50M_{\odot}$ for all waveforms for demonstration, as this is a commonly observed mass scale in LIGO observations. 
%\te{This choice seems pretty arbitrary. Looking at Fig 8 of 2111.03606, they seem to cluster closer to 60-70. This choice is important and should be more informed.}We will then examine the differences in optimization.

@@

We point out that our treatment to NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}. In the original calibration process, training waveforms are hybrid waveforms of NR and SpinAlignedEOB
(SEOB) waveforms. The low frequency inspiral part is taken from the SEOB
waveforms while the rest of the waveforms are taken from NR simulations.
Instead, we solely use NR waveforms for comparison, since we are only exploring the possibility of optimizing waveform models. Thus, for simplicity sake, we ignore this procedure. On the other hand, most NR waveforms used (for both training and testing) have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, in
which they are long enough to contain part of the inspiral segment and all
merger and ringdown frequency information. We take the frequency limits as
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set higher
than that in the original calibration due to NR length. When compared with
IMRPhenomC, the frequency range is slightly extended to have a higher maximum
frequency. We have the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$, we see the validation loss in 
Fig.~\ref{fig:loss} plateau at around $N=12000$, hence we end the optimization here.  
%\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

@@

To evaluate how well the optimization procedure generalize to waveforms that are
not in the training set, we evaluate the mismatch between the fine-tuned model
and the data provided in the SXS catalog for 536 NR waveforms. We select
waveforms that share the same part of the parameter space with the training
waveform, i.e. the waveforms with negligible eccentricity (${e<2\times10^{-3}}$)
and precession (${\chi_{x,y}<5\times10^{-3}}$). Fig.~\ref{fig:intrin_space}
shows how the training and testing waveforms are distributed in the
$q-\chi_{PN}$ space.

To illustrate the effect of optimiziation on an individual waveform level, we
compare the mismatch of a particular waveform before and after optimization with
respect to the NR waveform taken directly from the SXS catalog in Fig.
\ref{fig:0154}. Compared to the original IMRPhenomD waveform, the
optimized waveform has smaller residual from NR waveform both in amplitude and
phase, particularly in the inspiral region, where the amplitude displays a
$50\%$ reduction in error. For a fair comparison, we selected one of the testing
waveforms from the catalog presented in \citep{khan2016frequency}.

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire training dataset. Fig. \ref{fig:q148} shows the
distribution of log-mismatch for the testing waveforms before and after the
optimization procedure. For generality, we use a constant PSD in our loss
function. One can see the distribution have been skewed toward lower mismatch,
where the peak of the distribution is shifted by approximately an order of
magnitude, and a median mismatch is reduced by 50\%. When using
$\mathcal{L}_{\mathrm{fl}}$, we observe a similar improvement with a 22.9\%
decrease in the median.

@@

Note that the IMRPhenomD model was initially constructed and fitted using the
{\zdethp} weighted mismatch. To ensure a fair comparison, we apply the same
methods using the {\zdethp} PSD exhibit superior improvement compared to the
unweighted mismatch. We find no significant qualitative and quantitative
difference between the two PSDs in the resulting distribution of mismatches.

To understand whether additional training data can further improve the
performance of the model, we include waveforms that are not presence in the
original IMRPhenomD calibration in our training dataset with parameters
tabulated in Tab. \ref{tab:q1248}. We specifically choose to use $q=2$ events
since we have abundant $q=2$ NR waveforms to validate the final result. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the optimized
distribution remains comparable to the original distribution, indicating that
the original dataset is sufficient for this task. Similarly, utilizing the
{\zdethp} PSD to optimize the loss function with additional waveforms results in 
similar level of improvement.

To investigate the performance of recalibration over the source parameter space,
we plot the improvement of mismatch in log projected over the parameter space of
$q$ vs. $\chi_{\mathrm{PN}}$ in Fig. \ref{fig:ps_q148_qchi}. On the mass ratio
axis, we can see the waveforms with $q\le4$ show most consistent average
improvement. This is because that part of the parameter space is better covered
by the training waveform, as compared to testing waveforms from $q=4$ to $q=8$,
where some testing waveforms lie outside the source parameter space covered by
the training set of waveforms. On the spin axis, we can see the waveform with
$\chi_{\textrm{PN}}$ close to zero show the most consistent improvement, and we
start deviate from 0, there could be fluctuations in the average improvement. On
top of that, in the $q\le4$ region, there seems to be more consistent
improvement of waveforms with $\chi_{\textrm{PN}}<0$. This can be understood as
since the original waveform was developed for aligned spin system, the waveforms
with $\chi_{\textrm{PN}}$ is less well-fitted \kw{Varify this.}, hence there
could be more room for improvement. To show this, we plot the parameter space of
$\chi_1$ vs. $\chi_2$ in Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal
axis, i.e. $\chi_1\approx\chi_2$, show good mismatch improvements as discussed
above. Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right
($\chi_1>0>\chi_2$) regions respond to optimization differently. In the top-left
region, waveforms generally improve slightly with along optimization. However,
waveforms in the bottom-right region do not improve after optimization. Some
waveforms even turned worse after optimization.

 Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying source parameters, it is conceivable that different parts of the
source parameter space may not share the same set of optimal IMRPhenomD
parameters, mean there could be some trade-offs in accuracy between different
parts of the parameter space. If this is the cause of the lack of improvement in
the high mismatch tail of the distribution, segmenting the parameter space into
smaller subspaces should alleviate this problem. On the other hand, if the
ansatz lacks the correct parameterized form to capture the NR waveforms'
behavior as a function of the source parameters, the results will always be
biased, and we should not expect any improvement, even if we segment the
parameter space during training.

@@

% Moved down from intro
% In particular, we find it is easier IMRPhenomD
% favors certain parts of the parameter space. This means IMRPhenomD introduces
% systematic bias to other GW analysis tasks. This showcases the flaws of the
% ansatz and allows us to have a deeper understanding of Phenom models.

>>>>

\begin{abstract}
    We present a simple and general method of recalibrating gravitational wave
    (GW) phenomenological waveform models jointly. By using {\jax} and
    {\ripple}, we can perform automatic differentiation to functions, which
    allows us to use gradient-based optimization methods to recalibrate waveform
    coefficients in IMRPhenomD model. This method reduces systematic bias
    previously introduced to the model and generally can improve waveform
    accuracy. With recalibrated coefficients, we found that the typical
    \textit{mismatch} has a $50\%$ decrease. Furthermore, we analyze the
    accuracy base on the waveform's intrinsic parameters. We found that waveform
    accuracy has significant dependence on black hole spin. Reduced spin
    approximation introduces degeneracy in spin, which prevented further
    improvement. We isolated regions in the parameter space that does not fit
    the waveform ansatz. These results allow us to understand more about how to
    develop newer phenomenological models. 
\end{abstract}

@@

In the future, the Laser Interferometer Gravitational-wave Observatory (LIGO)
will finish its maintenance and start observing new gravitational wave (GW)
results. This new O4 run is expected to double the rate of current binary black
hole (BBH) observations \citep{abbott2020prospects}. Additionally, the
sensitivity of interferometers will be increased to capture more details of GW.
Having instruments with higher sensitivity, GW models of equal or higher
accuracy than observations should be used to extract GW information. Otherwise,
the extracted information would be affected more by GW models instead of
interferometer sensitivity, resulting in a bottleneck in GW analyses. Although
GW models are accurate enough for current analyses, the accuracy of current
models will no longer suffice for future data analyses
\citep{purrer2020gravitational}. Hence, it is necessary for us to develop and
improve GW models.

% Waveform families and slight motivation of why we take Phenom model
Currently, three families of GW models are commonly used. They are the
effective-one-body (EOB) \citep{ossokine2020multipolar, cotesta2020frequency, taracchini2014effective}, Numerical Relativity
(NR) surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2} and phenomenological (Phenom) models
\citep{husa2016frequency, khan2016frequency, garcia2020multimode, pratten2021computationally}. 
EOB models are constructed by mapping two masses onto an
effective body under an effective metric; NR surrogate models construct
waveforms using combinations of NR waveforms; Phenom models are formulated using
specific ansatz and inspiral approximations. While EOB and NR surrogate models
give better waveform approximants, Phenom waveforms can be produced much faster,
hence it is used mostly in data analysis tasks that requires many waveform
generations. This advantage scales up in data analysis tasks such as matched
filtering and parameter estimation, where many waveforms are required in each
run. This motivates us to improve upon the current framework of Phenom models,
thus can retain the advantage of fast waveform generation while improving the
model's accuracy.

Automatic differentiation (AD) is a method to calculate derivatives of functions
up to machine precision. In traditional numerical calculations, derivatives are
usually obtained through numerical derivatives. Symbolic derivatives were
available but it was less efficient. Both methods were not viable in machine
learning, where back-propagation requires precise and rapid derivative
calculations. In {\python}, packages including \texttt{pytorch} \citep{pytorch},
\texttt{tensorflow} \citep{tensorflow2015-whitepaper}, etc. utilizes AD to train
machine learning models. AD's algorithm is intuitive in nature. Functions
defined are decomposed into tree structures of primitives, such as addition or
function evaluations. Since these operations are fundamental, they were saved as
pairs internally. Differentiation proceeds forward following the tree structure,
with the application of the chain rule in each step to evaluate its derivative.
Analytic derivatives of such operations are applied in each step and the desired
derivative can then be obtained by composing back the original function
according to the original structure. {\ripple} \citep{ripple} was a new
implementation of IMRPhenomD, one of the models within the Phenom family. It was first implemented
in \texttt{lalsuite} using \texttt{C}. In order to make use of AD, it was
rewritten using \jax, a {\python} package that supports AD. Using
{\ripple}, one can apply AD to GW models to obtain precise derivatives, thus
allowing one to freely use derivative-based algorithms to perform data analyses. 
% Need to cite references for AD

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their intrinsic
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of optimized waveforms with original waveforms. We also show how the
optimization result differs with waveforms of different intrinsic parameters. In
Sec.~\ref{sec:discussion}, we address the difference between our calibrating
procedure with \citep{khan2016frequency}. We also explain how reduced spin
parameterization affects the accuracy of the model.

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{Owen:1995tm,husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as the mean square error (MSE) between the
two waveforms.

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}
	\mathcal{L}_{\mathrm{mean}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \,,\\
	\mathcal{L}_{\mathrm{norm}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}}\,,
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behaviour during the optimization process.
The simple average, $\mathcal{L}_{\mathrm{mean}}$, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, $\mathcal{L}_{\mathrm{mean}}$ allows the loss function to automatically adjust
and preferencially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.

@@

To compute the loss functions, we have to choose NR waveforms for calculating the mismatch.
Originally, 19 NR simulations  were used to calibrate IMRPhenomD \citep{khan2016frequency}; 9 of these 
are from the SXS catalog \citep{boyle2019sxs} and 10 BAM simulations. 
As BAM waveforms are not publicly available, we cannot use a training set identical to the original work.
Instead, we take the same nine waveforms from the SXS catalog plus two additional waveforms which closely mimic
two of the low mass ratio BAM simulations. 
The remaining BAM simulations have no close SXS counterparts, and are therefore not included in the training set.
Our main results therefore uses 11 NR waveforms for calibration which are listed in Tab.~\ref{tab:q148}.
Additional NR waveforms which are used for further calibration are listed in Tab.~\ref{tab:q1248}.

The training set has a maximum mass ratio of eight due to the lack 
of high mass ratio simulations in the SXS catalog.
In fact, the SXS catalog only has NR waveforms with $q\leq10$.
Nevertheless, we are interested in the behavior of IMRPhenomD model with small $q$,
as most BBH events observed by LIGO and Virgo have $q\leq8$.

The SXS NR waveforms are given as time-series strain.
Since IMRPhenomD is modeled in the frequency domain, we need to Fourier transform the simulation results
in order to compute the mismatch, \eqref{eq:mismatch}.
For this, we taper the time-series using Tukey window~\citep{tukey}\footnote{
	Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown and $T$ is the duration of the
	entire GW strain. \te{What is $t_{\mathrm{RD}}$? and how is it computed?}
} before using standard FFT routines to compute the Fourier transform.

In addition to choosing the NR waveforms for the training set,
one needs to choose a relevant PSD for mismatch.
We have opted to use a flat PSD for the mismatch calculation,
as it provides results that are independent of the detector sensitivity and mass scale.
The use of a flat PSD ensures that the improvement in accuracy is due mainly
to the difference in high-dimensional fitting.
Additionally, we are interested in examining the effect of introducing a detector PSD
on the optimization process. For this, we have chosen the
zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}.
Since the total mass of the system scales with the frequency of the waveform,
we must choose a corresponding mass scale to match the frequency range of our noise PSD.
To demonstrate the effect of introducing a detector specific PSD,
we selected an arbitrary mass scale of $M=50M_{\odot}$, as binaries
of this mass are commonly observed by the LIGO and Virgo detectors~\citep{gwtc1, gwtc2, gwtc21, gwtc3}.
%\te{This choice seems pretty arbitrary. Looking at Fig 8 of 2111.03606, they seem to cluster closer to 60-70. This choice is important and should be more informed.}We will then examine the differences in optimization.

@@

We point out that our treatment of NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}.
In the original calibration process, the training waveforms are hybrids of NR and SpinAlignedEOB (SEOB) waveforms.
The low frequency inspiral part is taken from the SEOB
waveforms while the rest is taken from NR simulations.
Instead, we solely use NR waveforms for calibration since most NR waveforms used (for both training and testing)
have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, to contain part of the inspiral segment and all
merger and ringdown frequency information. We use the frequency limits
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set slightly higher
than in the original calibration due to the NR simulation length. 
We use the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of the GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency} because they likely lie in the
neighborhood of the minimum that we wish to find. 
We fix our learning rate, $\alpha$, to be $=10^{-6}$, which is small enough to ensure we 
don't move far from the minimum.
Finally, we stop the optimization when the validation loss stops decreasing.
This can be seen in Fig.~\ref{fig:loss} at around $12000$ iterations.

@@

To evaluate the effectiveness of the optimization, we analyze 536 NR waveforms 
from the SXS catalog in this study. We specifically select waveforms with 
negligible eccentricity (${e<2\times10^{-3}}$) and precession 
(${\chi_{x,y}<5\times10^{-3}}$) that are consistent with the constraints of 
the waveform model. As shown in Fig.~\ref{fig:intrin_space}, the intrinsic 
parameters of the chosen test waveforms fall within the parameter space defined 
by the training waveforms. This suggests that these test waveforms can be used 
for direct comparison with the original model.

To illustrate the improvement on an individual waveform level, we compare a
particular waveform before and after optimization in Fig. \ref{fig:0154}.
Compared to the original IMRPhenomD waveform, we can see the optimized waveform
has smaller residual from NR waveform both in amplitude and phase, particularly
in the inspiral region, where the amplitude displays a $50\%$ reduction in
error. For a fair comparison, we selected one of the testing waveforms from the
catalog presented in \citep{khan2016frequency}.

To quantify the effect of optimization over the entire dataset, We evaluate the
mismatch of all testing waveforms using a constant PSD weighted loss function,
$\mathcal{L}_{\mathrm{mean}}$. We present the resulting distribution in Fig.
\ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{norm}}$, we observe a similar
improvement with a 22.9\% decrease in the median. Even though the distribution
lacks a clear peak due to the problem discussed in Section
\ref{subsec:optimization}, both distributions show observable improvement.

@@

%\kw{Which plot is this paragraph referring to?}
Applying the same methods, we will see that the distributions of mismatches
calculated using the {\zdethp} PSD exhibit superior improvement compared to the 
unweighted mismatch.
The shape of the distribution is similar to that shown in Fig. \ref{fig:q148}.
This result is expected, as the IMRPhenomD model was initially constructed and
fitted using the {\zdethp} weighted mismatch. Consequently, the model is
anticipated to closely fit the NR waveforms while incorporating the influence of
the {\zdethp} PSD, rather than a constant PSD.

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248}. We 
specifically choose to use $q=2$ events since we have abundant $q=2$ NR waveforms 
to validate the final result. The new set of coefficients generated from this
optimization process yields only marginal improvements in the newly produced
waveforms, as seen in Fig. \ref{fig:q148_q1248_compare}. The high mismatch tail
of the optimized distribution remains comparable in length and endpoint to the
original distribution, indicating that our procedure is incapable of improving
these waveforms. Similarly, utilizing the {\zdethp} PSD to optimize the loss
function with additional waveforms leads to similar improvement in the resulting
distribution.

% We believe that either the ansatz does not suit waveforms within certain
% regions of parameter space or the set of optimized coefficients falls to a
% minimum where they do not fit some testing waveforms. If in some region of
% parameter space, the ansatz does not fit well with NR waveforms, it would
% never be able to have consistent improvement under optimization. Instead, it
% would surf around the loss manifold with random fluctuations in mismatches. If
% the set of coefficients does not fit, then dividing the parameter space into
% regions and fitting separate sets of coefficients should give better results.
Although most of the waveforms show improvement in figure \ref{fig:q148}, the
high mismatch tail of the distribution remains unaffected. Given that the
waveform model's ansatz may not be entirely compatible with NR, and the
optimization procedure is carried out over a distribution of waveforms with
varying intrinsic parameters, it is conceivable that some trade-offs in accuracy
exist between different parts of the parameter space. If this is the cause of
the lack of improvement in the high mismatch tail of the distribution,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right ($\chi_1>0>\chi_2$) 
regions respond to optimization differently. In the top-left 
region, waveforms generally improve slightly with along optimization. However, waveforms 
in the bottom-right region do not improve after optimization. Some waveforms even 
turned worse after optimization.

@@
